### 2024-07-24 reports/paper/main.tex
paper: Initial Arxiv template

% Arxiv styling
% Based on https://github.com/kourgeorge/arxiv-style
\documentclass{article}
\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{floatrow}
%\usepackage{natbib}
\usepackage{doi}
\usepackage{graphicx}
\usepackage{subcaption}

% Document
\begin{document}

%% Title
\title{PRELIMINARY. Leaf compression for tiny Random Forests}

%% Author info
\author{
    Jon Nordby \\
	Soundsensing AS \\
	\texttt{jon@soundsensing.no} \\	
}

% Remove the date
\date{}

\maketitle
\renewcommand{\abstractname}{\vspace{-\baselineskip}} % erase the space between authors and abstract

%% Abstract
\begin{abstract}	\noindent

% Keywords. Maximum 5
\noindent \textbf{Keywords}:

\end{abstract}

\section{Introduction}

\noindent

\newpage
\section{Background}

\subsection{Related work}

\section{Methods}

\newpage
\section{Results}

%% DISCUSSION
\section{Discussion}

\subsection{Further work}

%\newpage
\section{Conclusions}

\section*{Acknowledgements}
\noindent

% References
\newpage
\bibliographystyle{unsrt}
\bibliography{references.bib}

\end{document}

### 2024-07-24 src/data/openml/download.py
download: Make output directory if needed

@@ -15,7 +15,9 @@ def download_openml_cc18(out_dir):
     Takes around 400 MB on disk in total.
     """
 
-    # FIXME: create directories as needed
+    if not os.path.exists(out_dir):
+        os.makedirs(out_dir)
+
     tasks_path = os.path.join(out_dir, 'tasks.csv')
 
     if os.path.exists(tasks_path):

### 2024-07-24 src/data/openml/download.py
download: Make datasets subdir also

@@ -15,8 +15,9 @@ def download_openml_cc18(out_dir):
     Takes around 400 MB on disk in total.
     """
 
-    if not os.path.exists(out_dir):
-        os.makedirs(out_dir)
+    datasets_dir = os.path.join(out_dir, 'datasets')
+    if not os.path.exists(datasets_dir):
+        os.makedirs(datasets_dir)
 
     tasks_path = os.path.join(out_dir, 'tasks.csv')
 
@@ -55,7 +56,7 @@ def download_openml_cc18(out_dir):
         data = X.copy()
         data['__target'] = y
 
-        data.to_parquet(os.path.join(out_dir, 'datasets', f'{dataset_id}.parquet'))
+        data.to_parquet(os.path.join(datasets_dir, f'{dataset_id}.parquet'))
 
 
 if __name__ == '__main__':

### 2024-07-24 reports/paper/main.tex

@@
 %% Title
-\title{PRELIMINARY. Leaf compression for tiny Random Forests}
+\title{(preliminary title) Leaf compression for tiny Random Forests}
 
@@
 % Keywords. Maximum 5
-\noindent \textbf{Keywords}:
+\noindent \textbf{Keywords}: Random Forest, TinyML, model compression, tree-based ensemble
 
@@
 \section{Introduction}
+
+Big picture motivation.
 
@@
 
+TinyML setting. On device inference. Low power. Wireless sensing.
+Success cases for Random Forest in TinyML.
+Human Activity Recognition. \cite{elsts_are_2021}
+Animal behavior. \cite{tatler_high_2018} \cite{kleanthous_feature_2020} \cite{tran_iot-based_2022}
+Battery monitoring. \cite{mawonou_state--health_2021}
+
+Structure of a random forest.
+Decision nodes, leaf nodes.
+Voting rules. argmax / Hard majority.
+Probability average. Soft.
+
+Fully grown versus partially grown trees.
+
 \subsection{Related work}
+
+Optimization techiques for decision trees and decision tree ensembles like Random Forest.
+Integer quantization of features.
+FPGA/hardware related.
+SIMD tree evaluation.
+
+Optimization for microcontrollers \cite{tabanelli_optimizing_2022}
+Feature selection. \cite{uddin_guided_2015}, \cite{elsts_energy-efficient_2020}
+Early stopping. \cite{daghero_low-overhead_2022} \cite{daghero_adaptive_2021}
+Parallelization.
+Cache-awareness.
+
+Software tools for deploying RF to TinyML. Open-source
 
@@
 
+What we cover: Feature quantization. Leaf quantization. Leaf clustering.
+
+OpenML CC18 datasets.
+Selected / eliminated datasets.
+TODO: list exact identifiers for that which was eliminated
+
+Train a general model. Using scikit-learn.
+
+Apply compression strategy to trained model. Re-running predictions on test set.
+
+Cross validation.
+
+? Demonstrator of performance on one or few TinyML datasets.
+PAMAP.
+Comparisons with other frameworks.
+
+\footnote{The code for reproducing the experiments can be found at https://github.com/jonnor/leaf-quantization}
+
 \newpage
 \section{Results}
+
+Performance with 16bit and 8 bit feature quantization, vs floating point.
+
+Percentage of model space (bytes) used for leaves.
+Histogram over models from all the datasets. Using 16 bit 
+For float (no quantization), uint16 and uint8 quantization (with deduplication).
+Determines the saving potential we have.
+
+? claim. Model size is the typical main contraint for RF in TinyML settings. Execution time is less of a problem
 
@@
 \subsection{Further work}
+
+Leaf clustering for regression.
 
@@
 
+Martin Stensgård
+
 % References

### 2024-07-25 src/data/openml/download.py
download: Print relevant columns for dropped datasets

@@ -34,12 +34,13 @@ def download_openml_cc18(out_dir):
         # Remove tasks/datasets with missing values
         tasks_missing_values = tasks['NumberOfInstancesWithMissingValues'] > 0
         print('Dropping tasks with missing values')
-        print(tasks[tasks_missing_values])
+        print(tasks[tasks_missing_values][['tid', 'did', 'name']])
         tasks = tasks[~tasks_missing_values]
 
         tasks.to_csv(tasks_path)
 
     for dataset_id in tasks['did']:
+        # TODO: add retrying with delay, sometimes there are connection problems
 
         # This is done based on the dataset ID.
         dataset = openml.datasets.get_dataset(dataset_id,

### 2024-07-25 src/data/openml/download.py
download: Ignore datasets with over 255 features
Primarily image recognition, not relevant for RF without feature engineering

@@ -4,6 +4,10 @@ import os
 import openml
 import pandas
 import openml.datasets
+import structlog
+
+
+log = structlog.get_logger()
 
 
 def download_openml_cc18(out_dir):
@@ -33,11 +37,14 @@ def download_openml_cc18(out_dir):
 
         # Remove tasks/datasets with missing values
         tasks_missing_values = tasks['NumberOfInstancesWithMissingValues'] > 0
-        print('Dropping tasks with missing values')
-        print(tasks[tasks_missing_values][['tid', 'did', 'name']])
-        tasks = tasks[~tasks_missing_values]
+        tasks_too_many_features = tasks['NumberOfFeatures'] >= 255
+        drop_tasks = tasks_missing_values | tasks_too_many_features
+        print('Dropping tasks not fitting criteria')
+        print(tasks[drop_tasks][['tid', 'did', 'name']])
+        tasks = tasks[~drop_tasks]
 
         tasks.to_csv(tasks_path)
+        log.info('task-list-downloaded', tasks=len(tasks), path=tasks_path)
 
     for dataset_id in tasks['did']:
         # TODO: add retrying with delay, sometimes there are connection problems

### 2024-07-25 src/experiments/run.py
experiments: Run with scikit-learn defaults for a baseline

@@ -212,49 +212,7 @@ def optimize(estimator, n_samples, n_classes, leaf_quantization=None, leaves_per
 
             for i in range(len(e.tree_.value)):
                 e.tree_.value[i] = v[i]
-
-
-
-class CustomRandomForestClassifier(BaseEstimator, ClassifierMixin):
-
-    def __init__(self, clusters=None, leaf_quantization=None, **kwargs):
-        #print('INIT', leaf_quantization)
-
-        self.estimator = RandomForestClassifier(**kwargs)
-        self.clusters = clusters
-        self.leaf_quantization = leaf_quantization 
-
-    def get_params(self, deep=True):
-        params = self.estimator.get_params()
-        params['clusters'] = self.clusters
-        params['leaf_quantization'] = self.leaf_quantization
-        return params
-
-    def set_params(self, **parameters):
-        #print("SET", **parameters)
-        our_keys = set(['clusters', 'leaf_quantization'])
-        our_params = { k: v for k, v in parameters if k in our_keys }
-        rf_params = { k: v for k, v in parameters if k not in our_keys }
-        ret = self.estimator.set_params(**rf_params)
-
-        for k, v in our_params:
-            setattr(self, k, v)
-
-        return ret 
-
-    def predict(self, X):
-        return self.estimator.predict(X)
-
-    def predict_proba(self, X):
-        return self.estimator.predict_proba(X)
-
-    def fit(self, X, y):
-        ret = self.estimator.fit(X, y)
-        self.classes_ = self.estimator.classes_
-
-        optimize(self, leaf_quantization=self.leaf_quantization, clusters=self.clusters)
-
-        return ret        
+    
 
 def setup_data_pipeline(data, quantizer=None):
 
@@ -390,7 +348,7 @@ def run_datasets(pipeline, out_dir, run_id, quantizer=None, kvs={}, dataset_dir=
         dataset_dir = 'data/raw/openml-cc18/datasets'
 
     matches = glob.glob('*.parquet', root_dir=dataset_dir)
-    assert len(matches) == 63, len(matches)
+    assert len(matches) == 52, len(matches)
     for no, f in enumerate(matches):
         dataset_id = os.path.splitext(f)[0]
         dataset_path = os.path.join(dataset_dir, f)
@@ -469,12 +427,18 @@ def main():
     
     
     experiments = {
[…]

### 2024-07-25 src/experiments/run.py
experiments: Prep running optimized

@@ -372,72 +372,22 @@ def run_datasets(pipeline, out_dir, run_id, quantizer=None, kvs={}, dataset_dir=
 
 
 def main():
-
-    # optimization steps. To be compared in experiments / ablated
-    # reduce number of features. To <255
-    # quantize features to int16
-    # reduce number of trees
-    # try replace nodes with k-means quantized versions
-
-    # TODO: run a hyperparameter search on each dataset.
-    # Find appropriate parameters, when no optimizations used
-    # for a particular set of trees. 10 is a good starting point?
-    # 1. Feature quantize 16 bit
-    # 2. Leaf quantize 8 bit
-    # 3. Leaf+feature quantize
-    #
-    # 4. Leaf reduce.
-    # TODO: specify  clusters as max leaves per class. 1,2,4,8,16,32,64,128
-
-    # Research questions
-    # A) how well does feature and leaf quantization work?
-    # Hypothesis: 16 bit feature is ~lossless. A 8 bit leaf probabilities is ~lossless
-    # Results indicate that 16 bit featurs is indeed lossless. Within the margins of error for experiment
-    # TODO: merge the int16 support to emlearn
-    #
-    # TODO: check 8 bit leaf probability quantization on CC-18
-
-    # TODO: set a few model size budgets. See how much performance one gets within that boundary
-    # uncontrained, 100kB, 10kB, 1kB
-
-    # B) how well does leaf clustering work?
-    # Hypothesis: Can reduce leaf size by 2-10 without ~zero loss in performance. Can reduce overall model size by 2-5x
-    # Preliminary results do indicate that up to 5x model size reduction is possible with low perf drops
-    # however there are a few outliers, where even 0.8 the original size causes drop in performance 
-
-
-    # C) how does code generation vs data structure compare, wrt size and inference time
-    # Hypothesis: datastructure "loadable" is smaller in size, but slower in inference time
-    # 
-
-    # D) what are limitating factors for practical models on microcontrollers
-    # Hypothesis: Model size is the primary bottleneck/constraint
-    # accelerometer. 10 FPS
-    # sound. 25 FPS
-    # images 1 FPS
-    # maybe do a "worst case" analysis of largesr ensembles that fit on a micro. 10kB, 100kB.
-    # Execution speed for max depths.
[…]

### 2024-07-25 src/data/openml/download.py
download: Allow specifying path, log progress

@@ -1,5 +1,6 @@
 
 import os
+import sys
 
 import openml
 import pandas
@@ -46,7 +47,7 @@ def download_openml_cc18(out_dir):
         tasks.to_csv(tasks_path)
         log.info('task-list-downloaded', tasks=len(tasks), path=tasks_path)
 
-    for dataset_id in tasks['did']:
+    for dataset_id, dataset_name in zip(tasks['did'], tasks['name']):
         # TODO: add retrying with delay, sometimes there are connection problems
 
         # This is done based on the dataset ID.
@@ -56,16 +57,21 @@ def download_openml_cc18(out_dir):
             download_features_meta_data=False,
         )
 
-        print(dataset.description[:100])
-
         target = dataset.default_target_attribute
         X, y, categorical_indicator, attribute_names = dataset.get_data(target=target)
 
         data = X.copy()
         data['__target'] = y
 
-        data.to_parquet(os.path.join(datasets_dir, f'{dataset_id}.parquet'))
+        dataset_path = os.path.join(datasets_dir, f'{dataset_id}.parquet')
+        data.to_parquet(dataset_path)
+        log.info('dataset-downloaded', id=dataset_id, name=dataset_name)
 
 
 if __name__ == '__main__':
-    download_openml_cc18('data/raw/openml-cc18')
+    default_path = 'data/raw/openml-cc18'
+    path = default_path
+    if len(sys.argv) > 1:
+        path = sys.argv[1]
+
+    download_openml_cc18(path)

### 2024-07-25 src/experiments/run.py
experiments: Run rest of min_samples_leaf on 10 trees

@@ -384,10 +384,10 @@ def main():
         #'trees20': dict(n_estimators=20),
         #'trees40': dict(n_estimators=40),
 
-        'minsamples10-1': dict(min_samples_leaf=0.1),
-        #'minsamples10-2': dict(min_samples_leaf=0.01),
-        #'minsamples10-3': dict(min_samples_leaf=0.001),
-        #'minsamples_defaults': dict(min_samples_leaf=1),
+        #'minsamples10-1': dict(min_samples_leaf=0.1),
+        'minsamples10-2': dict(min_samples_leaf=0.01),
+        'minsamples10-3': dict(min_samples_leaf=0.001),
+        'minsamples_defaults': dict(min_samples_leaf=1),
 
        #'rf10_float': dict(dtype=float, target_max=1000.0),   
        #'rf10_32bit': dict(dtype=numpy.int32, target_max=2**31-1),   

### 2024-07-25 src/experiments/run.py
experiments: Support specifying hyperparameters as envvars

@@ -370,32 +370,38 @@ def run_datasets(pipeline, out_dir, run_id, quantizer=None, kvs={}, dataset_dir=
         log.info('dataset-run-end', dataset=dataset_id, dataset_no=no, **kvs)
         print(score)
 
+def autoparse_number(s):
+    if '.' in s:
+        return float(s)
+    else:
+        return int(s)
+
+def config_number_list(var : str, default : str, delim=',') -> list[int]:
+
+    s = os.environ.get(var, default)
+    tok = s.split(delim)
+    values = [ autoparse_number(v.strip()) for v in tok if v.strip() ] 
+
+    print(tok, values)
+
+    return values
 
 def main():
     
     repetitions = int(os.environ.get('REPETITIONS', '3'))
     folds = int(os.environ.get('FOLDS', '5'))
+    trees = config_number_list('TREES', '100')
+    min_samples_leaf = config_number_list('MIN_SAMPLES_LEAF', '1')
 
-    experiments = {
-       #'sklearn_defaults': dict(dtype=None),
+    experiments = {}
+    for t in trees:
+        for l in min_samples_leaf:
+            name = f'tree-minsamplesleaf-{t}-{l}'
+            experiments[name] = dict(n_estimators=t, min_samples_leaf=l)
 
-        #'trees5': dict(n_estimators=5),
-        #'trees10': dict(n_estimators=10),
-        #'trees20': dict(n_estimators=20),
-        #'trees40': dict(n_estimators=40),
-
-        #'minsamples10-1': dict(min_samples_leaf=0.1),
-        'minsamples10-2': dict(min_samples_leaf=0.01),
-        'minsamples10-3': dict(min_samples_leaf=0.001),
-        'minsamples_defaults': dict(min_samples_leaf=1),
-
-       #'rf10_float': dict(dtype=float, target_max=1000.0),   
-       #'rf10_32bit': dict(dtype=numpy.int32, target_max=2**31-1),   
-       #'rf10_16bit': dict(dtype=numpy.int16, target_max=2**15-1),
-        #'rf10_12bit': dict(dtype=numpy.int16, target_max=2**12-1),
-        #'rf10_10bit': dict(dtype=numpy.int16, target_max=2**10-1),
-       #'rf10_8bit': dict(dtype=numpy.int8, target_max=127),
-    }
+    print('Experiments:')
+    for k, v in experiments.items():
+        print(k, v)
 
[…]

### 2024-07-25 src/data/openml/download.py
download: Check if data exists

@@ -50,6 +50,11 @@ def download_openml_cc18(out_dir):
     for dataset_id, dataset_name in zip(tasks['did'], tasks['name']):
         # TODO: add retrying with delay, sometimes there are connection problems
 
+        dataset_path = os.path.join(datasets_dir, f'{dataset_id}.parquet')
+        if os.path.exists(dataset_path):
+            log.info('dataset-exists', id=dataset_id, name=dataset_name)
+            continue
+
         # This is done based on the dataset ID.
         dataset = openml.datasets.get_dataset(dataset_id,
             download_data=False,
@@ -63,7 +68,6 @@ def download_openml_cc18(out_dir):
         data = X.copy()
         data['__target'] = y
 
-        dataset_path = os.path.join(datasets_dir, f'{dataset_id}.parquet')
         data.to_parquet(dataset_path)
         log.info('dataset-downloaded', id=dataset_id, name=dataset_name)
 

### 2024-07-26 src/experiments/analyze.py
experiments: Make sure to store hyperparameters with experiment

@@ -208,7 +208,7 @@ def enrich_results(df):
     leaf_bytes_per_class = leaf_bytes_per_class.fillna(value=4).astype(int)
 
     # FIXME: take the feature precision into account
-    decision_node_bytes = 2
+    decision_node_bytes = 8
 
     df = df.rename(columns={'test_leasize': 'test_leafsize'}) # Fixup typo
 

### 2024-07-26 src/experiments/run.py

@@ -433,7 +433,13 @@ def main():
 
         run_id = uuid.uuid4().hex.upper()[0:6] + f'_{experiment}'
 
-        run_datasets(p, quantizer=quantizer, optimizers=optimizers, kvs=dict(experiment=experiment),
+        kvs = {}
+        kvs.update(config)
+        kvs['experiment'] = experiment
+        kvs['folds'] = folds
+        kvs['repetitions'] = repetitions
+
+        run_datasets(p, quantizer=quantizer, optimizers=optimizers, kvs=kvs,
             out_dir=out_dir, run_id=run_id, repetitions=repetitions, cv=folds)
 
 

### 2024-07-26 src/experiments/run.py
experiment: Allow defining experiment name as envvar

@@ -392,11 +392,12 @@ def main():
     folds = int(os.environ.get('FOLDS', '5'))
     trees = config_number_list('TREES', '100')
     min_samples_leaf = config_number_list('MIN_SAMPLES_LEAF', '1')
+    experiment = os.environ.get('EXPERIMENT', 'tree-minsamplesleaf')
 
     experiments = {}
     for t in trees:
         for l in min_samples_leaf:
-            name = f'tree-minsamplesleaf-{t}-{l}'
+            name = f'{experiment}-{t}-{l}'
             experiments[name] = dict(n_estimators=t, min_samples_leaf=l)
 
     print('Experiments:')

### 2024-07-26 src/experiments/run.py
experiments: Include training score

@@ -301,6 +301,11 @@ def cross_validate(pipeline, X, Y,
             for metric, scorer in scoring.items():
                 res[f'test_{metric}'] = scorer(opt, X_test, Y_test)
 
+            for metric in ['roc_auc']:
+                scored = scoring[metric]
+                res[f'train_{metric}'] = scorer(opt, X_train, Y_train)
+
+
             dfs.append(res)
 
         return dfs 

### 2024-07-26 src/experiments/run.py
experiments: Support specifying feature type

@@ -398,12 +398,16 @@ def main():
     trees = config_number_list('TREES', '100')
     min_samples_leaf = config_number_list('MIN_SAMPLES_LEAF', '1')
     experiment = os.environ.get('EXPERIMENT', 'tree-minsamplesleaf')
+    feature_dtype = os.environ.get('FEATURE_DTYPE', None)
 
     experiments = {}
     for t in trees:
         for l in min_samples_leaf:
             name = f'{experiment}-{t}-{l}'
-            experiments[name] = dict(n_estimators=t, min_samples_leaf=l)
+            config = dict(n_estimators=t, min_samples_leaf=l, dtype=feature_dtype)
+            if feature_dtype == 'int16':
+                config['target_max'] = (2**15)-1
+            experiments[name] = config
 
     print('Experiments:')
     for k, v in experiments.items():

### 2024-07-26 src/experiments/run.py
experiments: Support max_depth
TODO: switch between this and min_samples_leaf

@@ -399,12 +399,14 @@ def main():
     min_samples_leaf = config_number_list('MIN_SAMPLES_LEAF', '1')
     experiment = os.environ.get('EXPERIMENT', 'tree-minsamplesleaf')
     feature_dtype = os.environ.get('FEATURE_DTYPE', None)
+    max_depth = config_number_list('MAX_DEPTH', '')
 
     experiments = {}
     for t in trees:
-        for l in min_samples_leaf:
-            name = f'{experiment}-{t}-{l}'
-            config = dict(n_estimators=t, min_samples_leaf=l, dtype=feature_dtype)
+
+        for d in max_depth:
+            name = f'{experiment}-{t}-{d}'
+            config = dict(n_estimators=t, max_depth=d, dtype=feature_dtype)
             if feature_dtype == 'int16':
                 config['target_max'] = (2**15)-1
             experiments[name] = config
@@ -437,7 +439,7 @@ def main():
         rf = RandomForestClassifier(
             n_estimators=config.get('n_estimators', 10),
             min_samples_leaf=config.get('min_samples_leaf', 1),
-            #max_features=0.33,
+            max_depth=config.get('max_depth', None),
         )
         p.append(rf)
 

### 2024-07-27 src/data/har/pamap2.py
data: Add loading code for Human Activity Detection datasets

@@ -1,0 +1,173 @@
+
+"""
+Data loading tools for the PAMAP2 Physical Activity Monitoring dataset
+https://archive.ics.uci.edu/dataset/231/pamap2+physical+activity+monitoring
+
+License: MIT
+Author: Jon Nordby
+"""
+
+import os
+
+import pandas
+
+def assert_data_correct(data):
+
+    # check dataset size
+    assert len(data) == 2872533, len(data)
+    assert len(data.columns) == (3*13)+2, len(data.columns)
+
+    # check index
+    assert data.index.names == ['subject', 'time']
+    assert data.dtypes['activity'] == 'category'
+
+    # check subjects
+    subjects = data.index.levels[0]
+    n_subjects = subjects.nunique()
+    assert n_subjects == 9, (n_subjects, list(subjects.unique()))
+
+    # check activities
+    activity_counts = data.activity.value_counts()
+    assert activity_counts['transient'] == 929661, activity_counts['transient']
+
+    
+
+def load_data(path) -> pandas.DataFrame:
+    """
+    Load all data in the dataset
+
+    NOTE: Takes some tens of seconds
+    """
+
+    # Prepare metadata
+    a = load_activities().items()
+    a = sorted(a, key=lambda kv: kv[0])
+    activity_categories = list(map(lambda kv: kv[1], a))
+
+    columns = load_column_names()
+
+    # Load the data
+    def load_one(path):      
+        df = pandas.read_table(path, header=None, sep='\s+', names=columns)
+        df['subject'] = os.path.splitext(os.path.basename(path))[0]
+
+        # drop invalid columns
+        invalid_columns = [ c for c in df.columns if '_invalid_' in c ]
+        df = df.drop(columns=invalid_columns)
+
+        return df
+
+    data_dir = os.path.join(path, 'Protocol')
+    assert os.path.exists(data_dir), data_dir
+
+    dfs = [ load_one(os.path.join(data_dir, name)) for name in os.listdir(data_dir) ]
+    data = pandas.concat(dfs)
+
+    # Use proper types
+    data['time'] = pandas.to_timedelta(data.time, unit='s')
+    data['subject'] = data.subject.astype('category')
+    data['activity'] = pandas.Categorical.from_codes(data.activity, categories=activity_categories)
+
+    # Set a reasonable index
[…]

### 2024-07-27 src/data/har/uci.py

@@ -1,0 +1,159 @@
+
+"""
+Dataset loading tools for "Smartphone based Recognition of Human Activities and Postural transitions"
+https://www.archive.ics.uci.edu/dataset/341/smartphone+based+recognition+of+human+activities+and+postural+transitions
+
+Uses the raw accelerometer/gyro data instead of the pre-computed features.
+
+License: MIT
+Author: Jon Nordby
+"""
+
+
+import os
+import glob
+
+import pandas
+import numpy
+
+def assert_data_correct(data):
+
+    # check dataset size
+    assert len(data) == 1122772, len(data)
+    assert len(data.columns) == 3+3+1, list(data.columns)
+
+    # check index
+    assert data.index.names == ['subject', 'experiment', 'time'], data.index.names
+    assert data.dtypes['activity'] == 'category'
+
+    # check subjects
+    subjects = data.index.levels[0]
+    n_subjects = subjects.nunique()
+    assert n_subjects == 30, (n_subjects, list(subjects.unique()))
+
+    # check experiments
+    experiments = data.index.levels[0].unique()
+    assert len(experiments) == 30, (len(experiments), list(experiments))
+
+    # check activities
+    activities = data.activity.unique()
+    assert len(activities) == 12+1, (len(activities), activities)
+
+    activity_counts = data.activity.value_counts(dropna=False)
+    assert activity_counts['SITTING'] == 126677, activity_counts['SITTING']
+
+    
+
+def load_data(path) -> pandas.DataFrame:
+    """
+    Load all raw IMU data and labels in the dataset
+
+    NOTE: Takes some tens of seconds
+    """
+
+    # Prepare metadata
+    activities_path = os.path.join(path, 'activity_labels.txt')
+    with open(activities_path, 'r') as f:
+        data = f.read()
+        lines = data.split('\n')
+        activities = {}
+        for l in lines:
+            tok = l.strip().split(' ')
+            if len(tok) == 2:
+                index, name = tok
+                activities[int(index)] = name
+
+    activities = pandas.Series(activities)
+    assert len(activities) == 12
+
+
+    # Load the data
+    raw_data_path = os.path.join(path, 'RawData')
+    def load_one(path, prefix):
+        path = os.path.join(raw_data_path, path)
+        samplerate = 50
+
+        # Extract metadata from filename
[…]

### 2024-07-27 reports/paper/main.tex

@@
 %% Title
-\title{(preliminary title) Leaf compression for tiny Random Forests}
+\title{(preliminary title) Model compression for Random Forests using leaf clustering}
 
@@
 \begin{abstract}	\noindent
+The Random Forest algorithm has been very successful, and continue to be a powerful choice on low-power microcontrollers with small amounts of memory and compute.
+Current size-optimized implementations of Random Forest tend to use hard majority voting, where each decision tree returns the most probable class, instead of the class proportions (probabilities).
+We show that the hard majority voting rule can lead to a considerable drop in predictive performance on some datasets.
+To find a better tradeoff between predictive performance and model size, we propose to use quantization and K-means clustering to adaptively select a small set of class proportions to store for the decision trees.
+We evaluate the proposed method on 60 classification datasets from the OpenML CC18 benchmark suite.
+Find that only a few class probabilities need to be stored to retain good performance,
+with small increase in model size over hard majority voting.
+When combined with 16-bit integer quantization of the input features, the method reached average model compression of XX, when staying within XXX percentage point performance drop.
+The method is implemented in the open-source TinyML library emlearn, and demonstrated running on microcontrollers for Human Activity recognition tasks.
 
@@
 Big picture motivation.
+TinyML use-cases.
+TinyML setting. On device inference. Low power. Wireless sensing.
 
@@
 
-TinyML setting. On device inference. Low power. Wireless sensing.
+\cite{random_forest_breiman2001}
+
 Success cases for Random Forest in TinyML.
@@
 Animal behavior. \cite{tatler_high_2018} \cite{kleanthous_feature_2020} \cite{tran_iot-based_2022}
-Battery monitoring. \cite{mawonou_state--health_2021}
 
+% REGRESSION, not classification Battery monitoring. \cite{mawonou_state--health_2021}
+
+TODO: illustration of tree structure
+TODO: illustration of the storage, of decision nodes and leaves
 Structure of a random forest.
@@
 Fully grown versus partially grown trees.
+Fully grown, all leaves with end up with class proportions.
+However when there are restrictions. Such as $max_depth$, $min_samples_leaf$, $min_samples_split$ etc.
+In resource constrained environments, this is the more realistic case.
+
+Class probabilities. 32 bit floating point.
+$leaf_size: n_leaves * n_classes * 4 bytes$
+
+Ensemble voting rule:
+\begin{verbatim}
+probabilities[n_trees, n_classes]
+for tree in forest:
+   tree_probabilities = predict_tree(tree)
+   probabilities[tree] = tree_probabilities
+proba = mean(probabilities, axis=1)
+\end{verbatim}
+
+As implemented in scikit-learn \cite{scikit-learn} RandomForestClassifier.
+
+Hard majority. argmax. With leaf-deduplication.
+Using 1 byte for the class indices, allows up to 255 classes.
+
+$leaf_size = n_classes * 1 * 1 byte$
+
+Voting rule:
+\begin{verbatim}
+votes[n_classes] = {0,...}
+for tree in forest:
+   class_index = predict_tree(tree)
+   votes[class_index] += 1
+out = argmax(votes)
+proba = votes / n_classes
+\end{verbatim}
+
+As implemented in emlearn 0.9.0 (2019) \cite{emlearn}.
+
+Proposed. Use soft voting rule. Reduce $n_leaves$ using clustering and reduce bitdepth using quantization.
 
@@
 
-Optimization techiques for decision trees and decision tree ensembles like Random Forest.
+Optimization techniques for decision trees and decision tree ensembles such as Random Forest has been studied extensively over the last decades, including in a low-power, low-compute setting.
+
 Integer quantization of features.
@@
 \section{Methods}
-
 What we cover: Feature quantization. Leaf quantization. Leaf clustering.
 
-OpenML CC18 datasets.
-Selected / eliminated datasets.
+OpenML CC18 datasets. \cite{oml-benchmarking-suites}
+To avoid needing data imputation, the datasets with missing values were ignored.
 TODO: list exact identifiers for that which was eliminated
 
-Train a general model. Using scikit-learn.
+Train a general model. Using scikit-learn. \cite{scikit-learn}
+Using 10 trees ({n_estimators}). And {min_samples_leaf=0.01}.
+NOTE: clustering will probably save more in percent, the more leaves are used.
+Increases both with the number of trees, and the effective depth
+- but total model sizes in excess of 100 kB is less relevant for the typical usecase
+
+TODO: consider other values for {min_samples_leaf} ?
+Either multiple, or tune on datasets.
+It will affect leaf growth.
+
+Metric. AUC ROC one-versus-all.
+
+Numerical variables are scaled using RobustScaler.
+Categorical variables are encoded using OrdinalEncoder.
 
@@
 
-Cross validation.
+Cross validation using stratified k-fold.
 
-? Demonstrator of performance on one or few TinyML datasets.
-PAMAP.
-Comparisons with other frameworks.
+? Demonstrator of performance on one or few TinyML tasks/datasets.
+HAR: PAMAP. UCI DSADS. OPPORTUNITY ? F1 scores. 
+Comparisons with other frameworks. Leaf cluster (emlearn), m2cgen, micromlgen
+Hardware architectures. ARM Cortex M0+ (no FPU). ESP32 (FPU). Alt: Cortex M4F
+TODO: visualization. Panel over HW,dataset of scatterplots with model size and inference time
+Need to implement feature computation also.
 
@@
 
-Performance with 16bit and 8 bit feature quantization, vs floating point.
+0. Baseline for comparison.
+1. Performance with 16bit and 8 bit feature quantization, vs floating point.
+Without any changes to the leaves.
+Practically no degradation in performance.
 
-Percentage of model space (bytes) used for leaves.
+INTERMEDIATE. Percentage of model space (bytes) used for leaves.
 Histogram over models from all the datasets. Using 16 bit 
 For float (no quantization), uint16 and uint8 quantization (with deduplication).
-Determines the saving potential we have.
+Determines the saving *potential* we have.
 
-? claim. Model size is the typical main contraint for RF in TinyML settings. Execution time is less of a problem
+REAL WORLD TEST.
+? Would have to make some illustrations of this.
+For example on HAR task/datasets, that on relevant microcontrollers, memory is problem before execution time. Need also to account for feature extraction time, which dominates.
+Typical workflow. Set hard constraints on model size. Optimize for maximal predictive performance. Say 1kB, 10kB, 100kB - or maybe steps of 2x
+VISUALIZATION. Model size vs predictive performance. And inference time vs predictive performance. Multiple model sizes. One line per framework, linked
 
@@
 
+? claim. Model size is the typical main constraint for RF in TinyML settings.
+Execution time (and power consumption) is less of a problem.
+
 \subsection{Further work}
 
-Leaf clustering for regression.
+Here we only considered the use of Random Forest for classification.
+The models are also used for regression.
+Explore whether leaf clustering could also be useful for regression.
 

### 2024-07-28 reports/paper/main.tex

@@
 
+\begin{figure}[h!]
+\begin{center}
+  \includegraphics[width=0.98\linewidth]{reports/figures/hyperparam-perfdrop-trees-strategies.png}
+  \end{center}
+  \caption{ Performance of relative to the strong baseline model, averaged over all datasets.
+With no restriction on tree growth ($min_samples_leaf=1$, leftmost side), majority voting and soft voting is identical.
+Majority voting performs worse compared to soft voting when tree growth is restricted, and this gets worse with lower number of trees.
+Soft voting benefits from tree growth restrictions, especially with lower number of trees.
+}
+  \label{fig:hyperparameter_strategies}
+\end{figure}
+
+Figure \ref{fig:leaf_proportions}).
+
 INTERMEDIATE. Percentage of model space (bytes) used for leaves.
@@
 Determines the saving *potential* we have.
+
+\begin{figure}[h!]
+\begin{center}
+  \includegraphics[width=2in]{reports/figures/leaf-uniqueness.png}
+  \end{center}
+  \caption{ More unique leaf values are produced when there are restrictions on tree growth. }
+  \label{fig:leaf_uniqueness}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+  \includegraphics[width=2in]{reports/figures/leaf-proportion.png}
+  \end{center}
+  \caption{The proportion of model size used by leaf nodes make up changes as restriction on tree growth. TODO: write more }
+  \label{fig:leaf_proportions}
+\end{figure}
 

### 2024-07-28 src/experiments/analyze.py
Compute pareto fronts and model size gains

@@ -201,14 +201,16 @@ def plot_size_improvement(df, path):
 
 
 
-def enrich_results(df):
+def enrich_results(df,
+        leaf_node_bytes_default=4,
+        decision_node_bytes_default = 8):
 
     # compute storage size
     leaf_bytes_per_class = df['leaf_bits'] / 8
-    leaf_bytes_per_class = leaf_bytes_per_class.fillna(value=4).astype(int)
+    leaf_bytes_per_class = leaf_bytes_per_class.fillna(value=leaf_node_bytes_default).astype(int)
 
     # FIXME: take the feature precision into account
-    decision_node_bytes = 8
+    decision_node_bytes = decision_node_bytes_default
 
     df = df.rename(columns={'test_leasize': 'test_leafsize'}) # Fixup typo
 

### 2024-07-28 reports/paper/main.tex

@@
 \begin{document}
+
+%% Custom commands
+% Inline reference to a hyperparameter.
+% Often contains underscores, which must be detokinized
+\newcommand{\hyperparam}[1]{\texttt{\detokenize{#1}}}
 
@@
 Fully grown, all leaves with end up with class proportions.
-However when there are restrictions. Such as $max_depth$, $min_samples_leaf$, $min_samples_split$ etc.
+However when there are restrictions. Such as \hyperparam{max_depth}, \hyperparam{min_samples_leaf}, \hyperparam{min_samples_split} etc.
 In resource constrained environments, this is the more realistic case.
@@
 0. Baseline for comparison.
+TODO: show a plot of the baseline performance?
+
 1. Performance with 16bit and 8 bit feature quantization, vs floating point.
@@
 Practically no degradation in performance.
+TODO: add some supporting references to 16 bits being enough. Both in general, and specific to decision trees / random forest
+
+\begin{figure}[h!]
+\begin{center}
+  \includegraphics[width=0.98\linewidth]{reports/figures/int16-vs-float.png}
+  \end{center}
+  \caption{ Performance of different feature representations, relative to the strong baseline model.
+Computed across all datasets and hyperparameters.
+16-bit integer quantized features shows practically identical performance to full precision floating point.
+}
+  \label{fig:int16_vs_float}
+\end{figure}
 
@@
   \end{center}
-  \caption{ Performance of relative to the strong baseline model, averaged over all datasets.
-With no restriction on tree growth ($min_samples_leaf=1$, leftmost side), majority voting and soft voting is identical.
-Majority voting performs worse compared to soft voting when tree growth is restricted, and this gets worse with lower number of trees.
+  \caption{ Performance relative to the strong baseline model, averaged over all datasets.
+With no restriction on tree growth (\hyperparam{min_samples_leaf=1}, leftmost side), majority voting and soft voting is identical.
+When tree growth is restricted, majority voting performs worse compared to soft voting, and this gets worse with lower number of trees.
 Soft voting benefits from tree growth restrictions, especially with lower number of trees.
@@
 
-Figure \ref{fig:leaf_proportions}).
+As the trees are more restricted, the uniqueness of leaf nodes goes up. 
+See Figure \ref{fig:leaf_uniqueness}).
+This happens because the algorithm is force to stop splitting a node before reaching a perfectly pure node (consisting of samples from a single class), and instead is left with a leaf consisting of a data-dependent proportion.
+This happens independently of the exact hyperparameter used to limit tree growth.
+% CLAIM. Not yet verified for all possibilities. Just min_samples_leaf and max_depth
 
-INTERMEDIATE. Percentage of model space (bytes) used for leaves.
-Histogram over models from all the datasets. Using 16 bit 
-For float (no quantization), uint16 and uint8 quantization (with deduplication).
-Determines the saving *potential* we have.
+% TODO: run experiments for all hyperparameters restricting tree depth
+% But probably no need to include all figures in the paper
+% scikit-learn RandomForest hyperparameters affecting tree depth and leaf nodes (number/split/composition)
+% direct restrictors on leaves
+% min_weight_fraction_leaf, min_samples_leaf, max_leaf_nodes
+% indirect restrictions
+% max_depth, min_samples_split, min_impurity_decrease, ccp_alpha
 
-\begin{figure}[h!]
-\begin{center}
+Since unique leaves cannot be deduplicated and must be stored individually, this leads to a larger proportion of the total model size taken up by leaves.
+This can be seen in Figure \ref{fig:leaf_proportions}).
+This leads to a potential for saving space by optimizing how leaves are stored.
+
+TODO: also show \hyperparam{max_depth}
+
+FIXME: check which leaf size is used. Should be 32-bit floating point for.
+And int16 features?
+
+\begin{figure}
+\centering
+\begin{minipage}{.5\textwidth}
   \includegraphics[width=2in]{reports/figures/leaf-uniqueness.png}
-  \end{center}
   \caption{ More unique leaf values are produced when there are restrictions on tree growth. }
   \label{fig:leaf_uniqueness}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=2in]{reports/figures/leaf-proportion.png}
+  \caption{The proportion of model size used by leaf nodes make up changes as restriction on tree growth. TODO: write more }
+  \label{fig:leaf_proportions}
+\end{minipage}
 \end{figure}
 
-\begin{figure}[h!]
-\begin{center}
-  \includegraphics[width=2in]{reports/figures/leaf-proportion.png}
-  \end{center}
-  \caption{The proportion of model size used by leaf nodes make up changes as restriction on tree growth. TODO: write more }
-  \label{fig:leaf_proportions}
-\end{figure}
+TODO: results with leaf clustering (and quantization) enabled
+Thesis: is pareto optimal in terms of predictive performance vs model size, when compared to majority and full soft voting
+? how to show that this holds across all (or majority) of datasets
+
+\subsection{Tests on microcontroller hardware}
 
@@
 Typical workflow. Set hard constraints on model size. Optimize for maximal predictive performance. Say 1kB, 10kB, 100kB - or maybe steps of 2x
-VISUALIZATION. Model size vs predictive performance. And inference time vs predictive performance. Multiple model sizes. One line per framework, linked
+
+VISUALIZATION. Model size vs predictive performance. And inference time vs predictive performance. Multiple model sizes. One line per framework, linked datapoints. Demonstrate Pareto optimality
 

### 2024-07-29 src/experiments/run.py
experiments: Support more different depth regularizer

@@ -391,22 +391,41 @@ def config_number_list(var : str, default : str, delim=',') -> list[int]:
 
     return values
 
+def get_depth_limiter():
+
+    depth_limiters = dict(
+        min_samples_leaf = config_number_list('MIN_SAMPLES_LEAF', ''),
+        min_samples_split = config_number_list('MIN_SAMPLES_SPLIT', ''),
+        max_depth = config_number_list('MAX_DEPTH', ''),
+        max_leaf_nodes = config_number_list('MAX_LEAF_NODES', ''),
+        min_impurity_decrease = config_number_list('MIN_IMPURITY_DECREASE', ''),
+    )
+
+    depth_limiter = None # tuple: (metric, value)
+    depth_limiters_configured = [ k for k, v in depth_limiters.items() if v ]
+    assert len(depth_limiters_configured) == 1, depth_limiters_configured
+    metric = depth_limiters_configured[0]
+    value = depth_limiters[metric]
+
+    return metric, value
+
 def main():
     
     repetitions = int(os.environ.get('REPETITIONS', '3'))
     folds = int(os.environ.get('FOLDS', '5'))
     trees = config_number_list('TREES', '100')
-    min_samples_leaf = config_number_list('MIN_SAMPLES_LEAF', '1')
     experiment = os.environ.get('EXPERIMENT', 'tree-minsamplesleaf')
     feature_dtype = os.environ.get('FEATURE_DTYPE', None)
-    max_depth = config_number_list('MAX_DEPTH', '')
+
+    depth_limiter_metric, depth_limiter_values = get_depth_limiter()
 
     experiments = {}
     for t in trees:
 
-        for d in max_depth:
-            name = f'{experiment}-{t}-{d}'
-            config = dict(n_estimators=t, max_depth=d, dtype=feature_dtype)
+        for limit in depth_limiter_values:
+            name = f'{experiment}-{t}-{limit}'
+            config = dict(n_estimators=t, dtype=feature_dtype)
+            config[depth_limiter_metric] = limit
             if feature_dtype == 'int16':
                 config['target_max'] = (2**15)-1
             experiments[name] = config

### 2024-07-30 src/experiments/run.py
run: Fix depth limiting hyperparameters actually not set

@@ -387,8 +387,6 @@ def config_number_list(var : str, default : str, delim=',') -> list[int]:
     tok = s.split(delim)
     values = [ autoparse_number(v.strip()) for v in tok if v.strip() ] 
 
-    print(tok, values)
-
     return values
 
 def get_depth_limiter():
@@ -454,12 +452,11 @@ def main():
             quantizer = Quantizer(dtype=config['dtype'],
                 max_quantile=0.001, out_max=config['target_max'])
 
+        rf_hyperparams_names = set(config) - set(['target_max', 'dtype'])
+        rf_hyperparams = { k: config[k] for k in rf_hyperparams_names }
+
         # classifier
-        rf = RandomForestClassifier(
-            n_estimators=config.get('n_estimators', 10),
-            min_samples_leaf=config.get('min_samples_leaf', 1),
-            max_depth=config.get('max_depth', None),
-        )
+        rf = RandomForestClassifier(**rf_hyperparams)
         p.append(rf)
 
         run_id = uuid.uuid4().hex.upper()[0:6] + f'_{experiment}'

### 2025-12-29 src/experiments/run.py
experiments: Fix FEATURE_DTYPE=float

@@ -447,7 +447,7 @@ def main():
 
         # feature quantization (optional)
         quantizer = None
-        if config.get('dtype'):
+        if config.get('dtype', 'float') != 'float':
             quantizer = Quantizer(dtype=config['dtype'],
                 max_quantile=0.001, out_max=config['target_max'])
 

### 2025-12-29 reports/paper/main.tex

@@
 \usepackage{microtype}      % microtypography
+\usepackage[strings]{underscore} % make sure _ in strings like \texttt works 
 \usepackage{floatrow}
@@
 \begin{abstract}	\noindent
-The Random Forest algorithm has been very successful, and continue to be a powerful choice on low-power microcontrollers with small amounts of memory and compute.
-Current size-optimized implementations of Random Forest tend to use hard majority voting, where each decision tree returns the most probable class, instead of the class proportions (probabilities).
-We show that the hard majority voting rule can lead to a considerable drop in predictive performance on some datasets.
-To find a better tradeoff between predictive performance and model size, we propose to use quantization and K-means clustering to adaptively select a small set of class proportions to store for the decision trees.
-We evaluate the proposed method on 60 classification datasets from the OpenML CC18 benchmark suite.
-Find that only a few class probabilities need to be stored to retain good performance,
-with small increase in model size over hard majority voting.
-When combined with 16-bit integer quantization of the input features, the method reached average model compression of XX, when staying within XXX percentage point performance drop.
-The method is implemented in the open-source TinyML library emlearn, and demonstrated running on microcontrollers for Human Activity recognition tasks.
+The Random Forest algorithm has been very successful, and continue to be a powerful choice on low-power embedded systems with small amounts of memory and compute.
+Current size-optimized implementations of Random Forest inference tend to use hard majority voting, where each decision tree returns the most probable class, instead of the class proportions (probabilities).
+We show that while compact, the hard majority voting rule can lead to a large drop in predictive performance on some datasets.
+To enable a better tradeoff between predictive performance and model size,
+we propose two methods to compress the model:
+using leaf quantization with deduplication, or leaf clustering with K-means.
+By evaluating using 60 classification datasets from the OpenML CC18 benchmark suite,
+we find that only a few class probabilities need to be stored to retain good performance, with small increase in model size over hard majority voting.
+When combined with 16-bit integer quantization of the input features, reducing the number of trees, the method reached average model compression of XXX, when staying within XXX percentage point performance drop of the uncompressed baseline.
+The method is implemented in the open-source TinyML library emlearn,
+and demonstrated running on microcontrollers for Human Activity recognition tasks.
 
@@
 
+Importance of small models.
+Microcontroller. Limited program storage (FLASH) and memory (SRAM).
+
 \noindent
@@
 
-\cite{random_forest_breiman2001}
+Random Forest model was originally introduced by \cite{random_forest_breiman2001}
+
+Decision tree ensembles have several properties that make them attractive
+for resource-constrained environments.
+Fast inference
+Easy to implement with integer-only arithmetic.
+High predictive performance with minimal hyperparameter tuning.
+Works well with small and medium sized datasets.
 
@@
 
-TODO: illustration of tree structure
-TODO: illustration of the storage, of decision nodes and leaves
 Structure of a random forest.
-Decision nodes, leaf nodes.
-Voting rules. argmax / Hard majority.
+
+Logically, a decision tree ensemble consists of a set of binary decision trees.
+Each tree has decision nodes that evaluate a set of if-else conditions,
+comparing one of the features of the input data to a learned threshold value. 
+At the leaves of the trees are the classification predictions for that tree.
+
+The leaves can either be a class index, in which case the ensemble prediction is done by counting the number of votes and outputting the majority.
+
 Probability average. Soft.
+% TODO: have some references to proportions, when it was introduced?
 
-Fully grown versus partially grown trees.
-Fully grown, all leaves with end up with class proportions.
-However when there are restrictions. Such as \hyperparam{max_depth}, \hyperparam{min_samples_leaf}, \hyperparam{min_samples_split} etc.
-In resource constrained environments, this is the more realistic case.
+% TODO: illustration of logical tree structure
+
+There are multiple ways of representing the decision tree ensemble at inference time.
+% TODO: some references to other approaches
+% As implemented in scikit-learn \cite{scikit-learn} RandomForestClassifier.
+
+Here we use an approach consisting of two tables:
+a table of decision nodes, plus a table of leaf nodes.
+
+Each decision node consists of feature index, threshold value, left child index and right child index.
+The left/right indices can either be an offset to a new decision node, or an offset into the table of leaf nodes (encoded as negative values).
+This memory layout is illustrated in Figure XXX: FIXME.
+% TODO: illustration of the storage, of decision nodes and leaves
+
+Each entry in the leaf node table has XXXX class proportions.
 
@@
 $leaf_size: n_leaves * n_classes * 4 bytes$
+
+If there are no restrictions on tree depth,
+all leaves end up only containing samples from a single class.
+In this circumstance, storing only the class index (instead of class proportions) preserves all the information in the tree.
+However, when there are restrictions on tree depth
+(for example with scikit-learn hyperparameters \hyperparam{max_depth}, \hyperparam{min_samples_leaf}, \hyperparam{min_samples_split}, et.c.)
+the leaves will containe a mix of class proportions.
+
+In resource constrained environments, this is the more realistic case.
 
@@
 \end{verbatim}
-
-As implemented in scikit-learn \cite{scikit-learn} RandomForestClassifier.
 
@@
 
-Proposed. Use soft voting rule. Reduce $n_leaves$ using clustering and reduce bitdepth using quantization.
+Proposed. Use soft voting rule.
+Reduce $n_leaves$ using clustering and reduce bitdepth using quantization.
 
@@
 
-Optimization techniques for decision trees and decision tree ensembles such as Random Forest has been studied extensively over the last decades, including in a low-power, low-compute setting.
+Optimization techniques for decision trees and decision tree ensembles (such as Random Forest) has been studied extensively over the last decades, including in a low-power and low-compute settings.
 
 Integer quantization of features.
-FPGA/hardware related.
+FPGA/hardware related. FIXME: references
+
 SIMD tree evaluation.
+TODO: cite QuickScorer/RAPIDSCORER, Fast Inference of Tree Ensembles on ARM Devices
 
@@
 Software tools for deploying RF to TinyML. Open-source
+% TODO: cite m2cgen, emlearn, etc
 
 \section{Methods}
+
 What we cover: Feature quantization. Leaf quantization. Leaf clustering.
 
-OpenML CC18 datasets. \cite{oml-benchmarking-suites}
-To avoid needing data imputation, the datasets with missing values were ignored.
-TODO: list exact identifiers for that which was eliminated
+\subsection{Comparison of leaf compression strategies on OpenML CC18}
+
+A subset of datasets from the OpenML CC18 benchmarking suite\cite{oml-benchmarking-suites} were used as a basis for the evaluation. 
+To avoid needing data imputation, the datasets with missing values were dropped,
+and to avoid datasets which require substantial feature selection or engineering, datasets more than FIXME:XX features were dropped.
+The selection removed FIXME:XXX datasets, leaving a total of FIXME:XXX datasets. The complete list can be found in the results.
+% MAYBE: list exact identifiers that which was eliminated?
+
+Numerical variables are scaled using RobustScaler, and categorical variables are encoded using OrdinalEncoder.
 
 Train a general model. Using scikit-learn. \cite{scikit-learn}
-Using 10 trees ({n_estimators}). And {min_samples_leaf=0.01}.
-NOTE: clustering will probably save more in percent, the more leaves are used.
-Increases both with the number of trees, and the effective depth
-- but total model sizes in excess of 100 kB is less relevant for the typical usecase
 
-TODO: consider other values for {min_samples_leaf} ?
-Either multiple, or tune on datasets.
-It will affect leaf growth.
+Apply compression strategy to trained model. Re-running predictions on test set.
+Cross validation using k-fold, with k=5.
 
@@
 
-Numerical variables are scaled using RobustScaler.
-Categorical variables are encoded using OrdinalEncoder.
+All experiments were done with int16 as feature representation,
+which reduces the size of decision nodes considerably compared to float32, while still allowing a sufficient range to give equivalent results on most datasets/tasks.
+% FIXME: references to existing papers
+We also found that int16 features resulted in the same performance as float32 on the selected datasets (not shown).
+The conversion from original feature data to 16-bit integers was done using robust min-max scaling between 0.01 percentile and 99.9 percentile.
 
-Apply compression strategy to trained model. Re-running predictions on test set.
+An overview of the different model compression strategies can be seen in Table \ref{fig:leaf_uniqueness}. 
+The Baseline configuration uses the default scikit-learn hyper-parameters (as of version FIXME), 100 trees (\texttt{n_estimators}) and no limit to tree depth (\texttt{min_samples_leaf=1}).
+All other configurations used 10 trees (more representative for models for resource constrained targets), and a range of tree depths were considered by trying \texttt{min_samples_leaf} values between 1,2,4,8,16,32,64,128.
+For Quantize we use between 1-8 bit quantization, and for ClusterQ8 consider 1,2,4,8,16,32 clusters per class.
 
-Cross validation using stratified k-fold.
+%NOTE: clustering will probably save more in percent, the more leaves are used.
+%Increases both with the number of trees, and the effective depth
+%- but total model sizes in excess of 100 kB is less relevant for the typical usecase
+
+\begin{table}[h]
+\centering
+\begin{tabular}{llllll}
+\hline
+\textbf{Strategy} & \textbf{Trees} & \textbf{Leaf representation} & \textbf{Deduplication} & \textbf{Clustering} & \textbf{Options} \\
+\hline
+Baseline & 100 & float32 & No & No  \\
+Reduced & 10 & float32 & No & No \\
+Majority & 10 & none & Yes & No \\
+Deduplicate & 10 & float32 & Yes & No \\
+Quantize & 10 & uint8 & Yes & No & 1-8 bit quantization \\
+ClusterQ8 & 10 & uint8 & Yes & Yes & 1,2,4,8,16,32 clusters per class \\
+% ClusterQ4 & 10 & 4+4 bits in uint8 & Yes & Yes \\
+\hline
+\end{tabular}
+\caption{Experimental configurations}
+\label{tab:models}
+\end{table}
+
+\subsection{Evaluation on hardware for Human Activity Recognition}
 
 ? Demonstrator of performance on one or few TinyML tasks/datasets.
-HAR: PAMAP. UCI DSADS. OPPORTUNITY ? F1 scores. 
+HAR: PAMAP. UCI DSADS. OPPORTUNITY ?
+F1 scores.
+
+Feature process. TODO: decide. MiniRocket ?? Any pre-selection?
+
 Comparisons with other frameworks. Leaf cluster (emlearn), m2cgen, micromlgen
-Hardware architectures. ARM Cortex M0+ (no FPU). ESP32 (FPU). Alt: Cortex M4F
-TODO: visualization. Panel over HW,dataset of scatterplots with model size and inference time
-Need to implement feature computation also.
+
+Hardware architectures.
+RP2040 ARM Cortex M0+ (no FPU). 
+Nordic NRF52840 ARM Cortex M4F
+ESP32 (FPU).
+
+TODO: visualization. Panel over HW, dataset of scatterplots with model size and inference time
 
 \footnote{The code for reproducing the experiments can be found at https://github.com/jonnor/leaf-quantization}
+% FIXME: double check link before submission
 
@@
 % But probably no need to include all figures in the paper
-% scikit-learn RandomForest hyperparameters affecting tree depth and leaf nodes (number/split/composition)
-% direct restrictors on leaves
-% min_weight_fraction_leaf, min_samples_leaf, max_leaf_nodes
-% indirect restrictions
-% max_depth, min_samples_split, min_impurity_decrease, ccp_alpha
 
@@
 
-FIXME: check which leaf size is used. Should be 32-bit floating point for.
+FIXME: Leaf size used. Should be 32-bit floating point for leaf probabilities.
 And int16 features?
@@
 TODO: results with leaf clustering (and quantization) enabled
-Thesis: is pareto optimal in terms of predictive performance vs model size, when compared to majority and full soft voting
-? how to show that this holds across all (or majority) of datasets
+Thesis: is Pareto optimal in terms of predictive performance vs model size, when compared to majority and full soft voting
 
@@
 
-Here we only considered the use of Random Forest for classification.
-The models are also used for regression.
-Explore whether leaf clustering could also be useful for regression.
+Here we only considered the use of leaf quantization and leaf clustering for classification, however we suspect that these methods would also be beneficial for  regression tasks.
+The methods may also be useful for gradient-boosted trees (such as XGBoost, LightGBM, etc).
 
@@
 
-Martin Stensgård
+We would like to thank Martin Stensgård for multiple insightful and productive discussions about low-level optimization of decision tree inference on microcontroller CPU arcitectures.
+
+Soundsensing has in the period 2022-2025 received funding from Research Council of Norway as part of the project EARONEDGE\cite{earonedge_nfr}.
 
